Deciding Bitvector Termination with Program Synthesis

### 2014-07-06 papers/termination/paper.tex
wordings

While the summary  in Figure~\ref{fig:handletable} supports the observation
that the majority of existing termination analyses are designed for linear
programs and linear ranking functions, it also identifies another common
assumption made by state-of-the-art termination provers: that bit-vector
semantics and integer semantics give rise to the same termination behaviour. 
Thus, most of the techniques treat bit-vectors and floats as mathematical
integers and reals,
respectively~\cite{DBLP:conf/pldi/CookPR06,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10}.

By assuming bit-vector semantics to be identical to integer semantics, these
techniques ignore the wrap-around behaviour caused by overflows, which can
be unsound.  In Section~\ref{sec:motivation}, we show that integers and
bit-vectors exhibit incomparable behaviours with respect to termination,
i.e.  programs that terminate for integers may \emph{not} terminate for
bit-vectors and vice-versa.  Thus, abstracting bit-vectors to integers may
give rise to {\em unsound} and {\em incomplete} analyses.
%
%design decisions
%which is rather surprising given that bit-vectors and floats are ubiquitous in computer systems.

@@

We propose a general framework that uniformly computes lexicographic,
non-linear ranking functions supported by non-linear inductive invariants
for loops with arbitrary guards and transitions over bit-vectors and floats. 
It also computes non-linear recurrence sets, which comprise non-termination
proofs.  Our encoding uniformly multiple loops and nested loops, without
having to enumerate lassos.
%
%Our technique is {\emph complete} and can handle programs with non-linear operations, e.g. logical and.
%--In our design, the termination problem becomes as hard as finding ranking functions, rather than as hard as
%checking the validity of a termination argument. 
%We combine the generation of ranking functions with the generation of invariants such that the necessary 
%supporting invariants and the ranking function are discovered at the same time.

For this purpose, we phrase the termination and non-termination problems as
second order satisfaction problems.  We identify a fragment of second-order
logic that is expressive enough to capture our formulation, which we call
the \emph{synthesis fragment}.  Subsequently, we propose a method for
checking the satisfiability of a formula in the synthesis fragment by
solving an isomorphic program synthesis problem
(Section~\ref{sec:synthesis}).  Our method is sound and semi-complete, i.e. 
if a pair of a ranking function and a supporting invariant exist it will be
found, but if the program does not terminate we may not find a recurrence
set.  However if the program's transition relation is deterministic, our
method is complete.

@@

%\item We propose a non-trivial formulation of the (non-)termination problem inside a second order logic fragment with only one quantifier alternation.

@@

Figure~\ref{fig:handletable} illustrates the most common assumptions made by existing termination analyses:
\begin{itemize}
\item[(i)] programs use only linear arithmetic.
\item[(ii)] terminating programs have termination arguments in linear arithmetic.
\item[(iii)] the semantics of bit-vectors and mathematical integers are equivalent.
\item[(iv)] the semantics of floats and mathematical reals are equivalent.
\end{itemize}

@@

%% One difference from previous approaches is that our method is fully semantic. As a consequence, 
%% we do not care about the CFG (we do not need to encode the CFG 
%% through program counter in the second order constraints).

@@

In the rest of this section, we show that the synthesis fragment 
is expressive enough to encode both termination and non-termination. %allow the formulation of both the termination and the non-termination problem. 
Following our formulation as second order satisfaction, %the framing of the halting problem as a program synthesis problem,
in Section~\ref{sec:synthesis} we present a solver for the synthesis fragment that models bit-accurate semantics.

@@

Note that different formulations for unconditional termination are possible.
We are aware of a proof rule based on transition invariants, i.e.  supersets
of the transition relation's transitive closure
\cite{DBLP:conf/pldi/GrebenshchikovLPR12}.  This formulation assumes that
the second order logic has a primitive predicate for disjunctive
well-foundedness.  Conversely, our formulation in Definition~\ref{def:UT}
does not require a transitive closure operator, nor a disjunctive
well-foundedness predicate.  \\

@@

%% ---  We begin with the program under analysis (the PUA).  From this we produce a specification (in second order logic)
%% encoding the termination of the PUA.  From the logical specification, we produce a program SYNTH.
%% SYNTH takes as input a proof and asserts that the proof is incorrect on at least one of the inputs being tracked so far.
%% Therefore, SYNTH is unsafe iff there is some proof that is correct for all of the tracked inputs.  Now we use testing
%% techniques to try to find a proof that causes SYNTH to crash.  The testing techniques we use are: model checking,
%% exhaustive search and genetic programming.  For the latter 2 techniques, we combine SYNTH with another program
%% (GP or EXPLICIT respectively) to create a program that searches for test vectors (proofs) that cause SYNTH
%% to crash.

@@

\begin{proof}
 Since the {\sc ExplicitSearch} routine enumerates all programs (as can be seen by induction on
 the program length $l$), it will eventually enumerate a program that meets the specification
 in whatever set of inputs are currently being tracked.  Since the first order theory is
 decidable, the query in {\sc Verif} will succeed for this program, causing the algorithm to terminate.
 The set of correct programs is therefore recursively enumerable and algorithm~\ref{alg:cegis}
 enumerates this set, so it is semi-complete.

>>>>

Figure~\ref{fig:handletable} summarises the related work with respect to the
restrictions they impose on the transition relations as well as the form of
the ranking functions computed.  While it supports the observation that the
majority of existing termination analyses are designed for linear programs
and linear ranking functions, it also highlights another simplifying
assumption made by most state-of-the-art termination provers: that
bit-vector semantics and integer semantics give rise to the same termination
behaviour.  Thus, most of the techniques treat bit-vectors and IEEE floats
as mathematical integers and reals,
respectively~\cite{DBLP:conf/pldi/CookPR06,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10}.

By assuming bit-vector semantics to be identical to integer semantics, these
techniques ignore the wrap-around behaviour caused by overflows, which can
be unsound.  In Section~\ref{sec:motivation}, we show that integers and
bit-vectors exhibit incomparable behaviours with respect to termination,
i.e.  programs that terminate for integers may \emph{not} terminate for
bit-vectors and vice versa.  Thus, abstracting bit-vectors to integers may
give rise to {\em unsound} and {\em incomplete} analyses.

@@

We propose a general framework that uniformly computes lexicographic,
non-linear ranking functions supported by non-linear inductive invariants
for loops with arbitrary guards and transitions over bit-vectors and floats. 
It also computes non-linear recurrence sets, which comprise non-termination
proofs.  Our encoding treats multiple loops and nested loops uniformly,
without having to enumerate lassos.
%
%Our technique is {\emph complete} and can handle programs with non-linear operations, e.g. logical and.
%--In our design, the termination problem becomes as hard as finding ranking functions, rather than as hard as
%checking the validity of a termination argument. 
%We combine the generation of ranking functions with the generation of invariants such that the necessary 
%supporting invariants and the ranking function are discovered at the same time.

 For this purpose, we phrase the termination and non-termination problems as
second-order satisfaction problems.  We identify a fragment of second-order
logic that is expressive enough to capture our formulation, which we call
the \emph{synthesis fragment}.  Subsequently, we propose a method for
checking the satisfiability of a formula in the synthesis fragment by
solving an isomorphic program synthesis problem
(Section~\ref{sec:synthesis}).  Our method is sound and semi-complete, i.e. 
if a pair of a ranking function and a supporting invariant exist it will be
found, but if the program does not terminate we may not find a recurrence
set.  However if the program's transition relation is deterministic, our
method is complete.

@@

%\item We propose a non-trivial formulation of the (non-)termination problem inside a
% second-order logic fragment with only one quantifier alternation.

@@

Figure~\ref{fig:handletable} illustrates the most common simplifying
assumptions made by existing termination analyses:
%
\begin{itemize}
\item[(i)] programs use only linear arithmetic.
\item[(ii)] terminating programs have termination arguments in linear arithmetic.
\item[(iii)] the semantics of bit-vectors and mathematical integers are equivalent.
\item[(iv)] the semantics of floats and mathematical reals are equivalent.
\end{itemize}

@@

%% One difference from previous approaches is that our method is fully semantic. As a consequence, 
%% we do not care about the CFG (we do not need to encode the CFG 
%% through program counter in the second-order constraints).

@@

In the rest of this section, we show that the synthesis fragment 
is expressive enough to encode both termination and non-termination. %allow the formulation of both the termination and the non-termination problem. 
Following our formulation as second-order satisfaction, %the framing of the halting problem as a program synthesis problem,
in Section~\ref{sec:synthesis} we present a solver for the synthesis fragment that models bit-accurate semantics.

@@

Note that different formulations for unconditional termination are possible.
We are aware of a proof rule based on transition invariants, i.e.  supersets
of the transition relation's transitive closure
\cite{DBLP:conf/pldi/GrebenshchikovLPR12}.  This formulation assumes that
the second-order logic has a primitive predicate for disjunctive
well-foundedness.  Conversely, our formulation in Definition~\ref{def:UT}
does not require a transitive closure operator, nor a disjunctive
well-foundedness predicate.  \\

@@

%% ---  We begin with the program under analysis (the PUA).  From this we produce a specification (in
%% second-order logic)
%% encoding the termination of the PUA.  From the logical specification, we produce a program SYNTH.
%% SYNTH takes as input a proof and asserts that the proof is incorrect on at least one of the inputs being tracked so far.
%% Therefore, SYNTH is unsafe iff there is some proof that is correct for all of the tracked inputs.  Now we use testing
%% techniques to try to find a proof that causes SYNTH to crash.  The testing techniques we use are: model checking,
%% exhaustive search and genetic programming.  For the latter 2 techniques, we combine SYNTH with another program
%% (GP or EXPLICIT respectively) to create a program that searches for test vectors (proofs) that cause SYNTH
%% to crash.

@@

\begin{proof}
 Since the {\sc ExplicitSearch} routine enumerates all programs (as can be seen by induction on
 the program length $l$), it will eventually enumerate a program that meets the specification
 in whatever set of inputs are currently being tracked.  Since the first-order theory is
 decidable, the query in {\sc Verif} will succeed for this program, causing the algorithm to terminate.
 The set of correct programs is therefore recursively enumerable and algorithm~\ref{alg:cegis}
 enumerates this set, so it is semi-complete.
